### 2018-09-11 Writing/disparity_review.tex
Updated shit sandwich

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: best practices for disparity analysis.

@@

\noindent{\Large \bf Analysing disparity in morphology data: best practice and future directions}

@@

\subsection{Is disparity equivalent to diversity?}
One obvious aspect when measuring biodiversity is the measurement of raw species diversity (at any scale; $\alpha$, $\beta$, $\gamma$).
Such biodiversity metric intuitively describes the amount of reproductive isolated populations in a system.
On the other hand, disparity aims at describing the amount of measured traits in a system (see below), in such a way, a high morphological disparity value is a measurement of a high diversity of morphologies.
The link between both disparity and diversity has been scrutinised under different scenarios with most conclusions supporting an effective decoupling between both [CITE the disparity/diversity papers].
This is intuitively understandable when comparing the both metrics for two contrasted groups of organisms, for example, rodents would display a high diversity (@@@ sp) with a relatively low disparity (a majority of rodents have the typical rat shape) whereas carnivores would display a relatively lower diversity (@@@ sp) with a higher disparity (with body shapes ranging from a weasel to an elephant seal).
However, it is important to note that in more nuanced groups, diversity and disparity can be link where a really low diversity would tend to yield a low disparity and vice versa.

%ii) the ensuing ideas on evolvability
2 - Disparity can be linked to evolvability? A morphological disparate group is likely to occupy more niches (but see the debate for the link between traits and ecological niches [CITES]) and more likely to be the result from character displacement and/or competition.
Also, more disparate groups are more likely to lead to disparate radiations as different ecological strategies get explored.

% OU -> Disparity can be many things

\section{A diversity of disparities}
% IN -> There are many different disparity
This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
However, as with any great power, it comes with great responsibility. 
With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.

\subsubsection{Morphological disparity}

\subsubsection{Functional disparity}
3.2- Functional disparity: seems to work at any scale and can be used to test convergence (species with similar trait combination can have similar functions) or the link between form and function (Diaz paper). However, the link between trait and function can sometimes be hard to grasp since traits can be linked to multiple functions and multiple functions can be linked to a single trait.
\subsubsection{Ecological disparity}
3.1- Ecological disparity: promising tool at a macro-evolutionary scales (but maybe less at micro-evolutionary scales). E.g. the concept of ecological guilds can be distinguished in the fossil record (carnivoran guild etc.) but it might be more difficult to grasp at a micro-scale level (CITE plant trait/niche papers).
\subsubsection{Phylogenetic disparity}

%However, do these disparities represent anything more than the disparity of those traits rather than the disparity of the thing?

%Maybe. Disparity traits need to be carefully selected in order to properly reflect the thing. Disparity is only a proxy to describe the pattern and does not necesserly make conclusions on the process (but see disparity pattern vs disparity process)

\section{Trait disparity}
% IN -> If it's just traits, can all of the traits be used?
Once one has a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for spec).ies delimitation.
Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below

\subsection{Are different traits equivalent?}
%     Can disparity analysis work with any kind of trait?
A common practice in morphological disparity analysis is to recycle data from other studies which can introduce additional problems.
One classic case is using ``cladistic'' data in palaeobiology.
These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all of the data.
For example, in a study of the disparity of dietary niches, removing non-dietary characters is a good approach.
However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g. limb traits can be used for both locomotion and feeding).

\subsection{Are different dataset equivalent?}
%     Do discrete and continuous trait reflect the same thing? If yes, what does that imply?
Morphological disparity analysis can be based on three main types of data:
\begin{enumerate}
    \item discrete data that can either describe a feature (often referred to as ``cladistic''; e.g. the presence or absence of legs, the colour of a tail, etc. [CITE]) or count the number of features (e.g. number of digits, etc. [CITE]);
    \item continuous data that can be either simple measurements (e.g. body length, etc. [CITE]) or more abstracted measures (e.g. geometric morphometrics, mathematical contour description, etc. [CITE]);
    \item any mixture of discrete and continuous data (e.g. ratios [CITE]).
\end{enumerate}
None of these types of data are fundamentally better but each will be more appropriate for specific questions.
For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometric analysis) or continuous measurements of bones may be appropriate.
On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question.
This again illustrates the importance of framing the question.
In the above example, if the question is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g. wingspan, aspect ratio, etc.).
Conversely if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g. digit length, integumentary system, etc.).

\subsection{Does the part represent the whole?}
%     Does the part represent the whole?
Melanie's work.

%Bonus
There are many other considerations when choosing the appropriate data to use. 
It is essential to ensure that disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data.
For example, are the observed differences between two groups related to factors unrelated to group designation such as sexual dimorphism?
Are the taxa/groups being compared fairly consistent and truly comparable?
In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated [String of morpho data problems cite].
It is important to ensure that any patterns detected are not merely reflections of these features of the data.

Ultimately, data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
% OU -> And how do these traits need to be analysed?

\section{Disparity is multidimensional}
% IN -> Disparity is multidimensional by essence
% Visualisation problem
\subsection{Visualising the multidimensional problem}
\label{visualisation}
It is important to remember that visual interpretations of multidimensional data can easily be misleading.
For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
Conversely, when two groups do not overlap on a two or three dimensional plot, this does not mean they never overlap in the multidimensional space.
Furthermore, multidimensional spaces are not necessary Euclidean.
Such non-Euclidean spaces often have non-intuitive properties, for example straight lines in some dimensions are not straight in the space and, even less intuitively, the distance between points A and B might not be equal to the distance between points B and A [CITE Sylvain].

@@

\subsection{[Box]To ordinate or not to ordinate; that is the (multidimensional) question?}
\label{box_ordination}
Reducing the dimensionality of data through ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g. selecting only the axes that describe up to 95\% of the variation).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions.
This has clear advantages for interpreting the results.
For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).

@@

Because of these points, we strongly recommend that data for multidimensional analyses is not automatically ordinated, and that careful consideration is given to whether the question can be answered without ordination.
\subsection{[Box] end}

%     What do we measure from these multidimensional spaces?
% OU -> And how do we test these things?

\section{Disparity hypotheses}
% IN -> So how do we go for testing hypotheses in disparity?
%     What are our null expectations?
The final stage of most disparity analyses is statistical testing of the disparity metrics, to determine whether changes thought time or between groups are statistically significant. 
Similar to the points above, the statistical test to apply depends on which question is being tackled with disparity analysis, i.e. which hypothesis is being tested.
It also requires consideration of the data, the ordination method used (or the absence thereof) and the metric used.

For example, one commonly used test in morphological analysis, is the non parametric permutation analysis of variance [NPANOVA or PERMANOVA; CITE Anderson + examples].
This test and its implementation is perfectly valid to test whether two groups have a different variance/covariance, but is based on the pairwise distances of the group elements.
This test will ignore any ordination (if the data was ordinated) and is not based on a unidimensional disparity metric.
In other words, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the question.

@@

%     How do we take phylogeny into account?

\subsection{Phylogeny}
For many disparity analyses of macroevolutionary questions, phylogeny is often taken into account, whether for including ancestral state estimates or to incorporate phylogenetic structure in the data.
For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
and therefore will not covered further in this review.

\subsubsection{Ancestral state estimation}
When looking at disparity in the past (e.g. disparity-through-time) ancestral state estimations are a common approach.
Ancestral state estimation can be performed in two ways:
\begin{enumerate}
\item pre-ordination: the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original dataset[CITE].
\item post-ordination: the estimation is done after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
\end{enumerate}
Pre-ordination ancestral state estimation can change the ordinated space's geometry (i.e. the relationship between the points not estimated) and implies longer computational times but, once the morphospace is defined, it will remain constant (i.e. its properties will not change).
Post ordination ancestral state estimation will not change the ordinated space's geometry and will be faster to compute, however, it will change the morphospace's properties such as the number of parameters (i.e. elements in the space - this can be problematic for statistical tests down the line) or the variance on each axis [CITE Graeme's paper to be?].

@@

Alternatively, ancestral state estimation can be used to provide realistic null expectations from the data.
In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes.
Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution.
%     How can we analyse it through time?

@@

Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.
% The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
% Each individual carnivore species occupies a discrete region of this space related to its values for these traits.

The initial informal definition of disparity was ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (REFS).
Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation (REF Katie et al's paper somewhere here). 
Here we use a broader definition and suggest that disparity is a pattern of variation.
Throughout this review, we aim to highlight the diversity of ``disparity'' analyses and how they can aid our understanding of the structure of biodiversity.

% \subsection{What \textit{is} disparity?} %TG: maybe useful for box or paraphrapsing summaries.
% What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
% The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
% Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation. 
% There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
% Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
% The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).

% Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
% In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
% Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).

\section{Best practices for disparity analysis}

This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
However, as with any great power, it comes with great responsibility. 
With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.

\section{What is the question?}
Why perform a disparity analysis?
This seems like it should be obvious, but we stress its crucial importance.
The last decade has seen a vast array of new methods being published and implemented in freely available software, making disparity analyses easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
However, this (consciously or not) pushes authors towards performing disparity analyses simply because of their ease of implementation, rather than stemming from a question driven approach.
Before rushing into a disparity analysis, one should consider some basic caveats.
For example, one major thing to consider is circularity: if the aim is to explore the variation in a dataset, disparity analysis is perfectly valid.
However, if this exploratory analysis reveals distinct groups, it would be circular to then test for differences among those groups as if this had been the hypothesis all along [CITE stats 1.0].
This can easily be avoided by using a subset of the data for preliminary data exploration, and then using the full dataset to test for differences among any groups that emerge.
Clearly framing questions and their null expectations will help in designing better disparity analyses.

\section{Collect the appropriate data}
Once one has a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for species delimitation.
Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).

\subsection{Data types}
Morphological disparity analysis can be based on three main types of data:
\begin{enumerate}
    \item discrete data that can either describe a feature (often referred to as ``cladistic''; e.g. the presence or absence of legs, the colour of a tail, etc. [CITE]) or count the number of features (e.g. number of digits, etc. [CITE]);
    \item continuous data that can be either simple measurements (e.g. body length, etc. [CITE]) or more abstracted measures (e.g. geometric morphometrics, mathematical contour description, etc. [CITE]);
    \item any mixture of discrete and continuous data (e.g. ratios [CITE]).
\end{enumerate}
None of these types of data are fundamentally better but each will be more appropriate for specific questions.
For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometric analysis) or continuous measurements of bones may be appropriate.
On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question.
This again illustrates the importance of framing the question.
In the above example, if the question is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g. wingspan, aspect ratio, etc.).
Conversely if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g. digit length, integumentary system, etc.).

\subsection{How much data should be collected?}
Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of specimens and traits.
\begin{itemize}
    \item \textit{Taxonomic scope.}
    Morphological disparity has been studied at both inter-[CITE] and intra-specific levels [CITE].
    Taxonomic scope will depend on the question being investigated, and what type of variation is being analysed.
    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation to capture the full variety of skull shapes, particularly that seen in domestic dogs (CITE).
    Conversely, if a study focuses on canid skull variation through time, interspecific data is most appropriate.
    % intra-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
    \item \textit{Sample size.}
    It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
    For example, if the aim is to compare disparity among groups, one should collect, where possible, a similar amount of data for each group.
    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction [CITE]).
    Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
    \end{itemize}

>>>>

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: disparate disparity analysis.

@@

\noindent{\Large \bf The disparite ways to analysis morphological disparity} %TG: oooooh the ugly title!

@@

\subsection{Is disparity equivalent to diversity?}
One obvious aspect when measuring biodiversity is the measurement of raw species diversity (at any scale; $\alpha$, $\beta$, $\gamma$).
Such biodiversity metric intuitively describes the amount of reproductive isolated populations in a system.
On the other hand, disparity aims at describing the amount of measured traits in a system (see below), in such a way, a high morphological disparity value is a measurement of a high diversity of morphologies (i.e. shapes or body plans).
The link between both disparity and diversity has been scrutinised under different scenarios with most conclusions supporting an effective decoupling between both [CITE the disparity/diversity papers].
This is intuitively understandable when comparing both metrics for two contrasted groups of organisms, for example, living rodents would display a high diversity ($>$2200 sp; CITE Wilson\&Reeder) with a relatively low disparity (a majority of rodents have the typical rat shape) whereas carnivores would display a relatively lower diversity ($>$280 sp; CITE Wilson\&Reeder) with a higher disparity (with body shapes ranging from a weasel to an elephant seal).
However, it is important to note that in more nuanced groups, diversity and disparity can be link where a really low diversity would tend to yield a low disparity and vice versa.

This decoupling has led researchers to work on the relation between disparity and evolvability: a more disparate group (population, group, etc...) will have more genetic diversity to adapt to biotic and abiotic changes [CITES].
This led to the use of disparity as a tool for investigating morphological modularity [CITE].
However, this is based on the assumption that there is a link between diversity in traits (disparity) and subsequent diversity in genes and/or in ecological niches ([Trait plant paper CITE] see below for more details).
Ultimately, both links between disparity and diversity and disparity and evolvability are dependent on the multiple definitions of disparity and the multiple datasets used.

\section{A diversity of disparities}
% IN -> There are many different disparity
This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
From a same analysis comparing morphological traits among different organisms, one can study four main types of disparity:

\begin{enumerate}

\item{Morphological disparity.}
The disparity of morphological trait configurations has been at the genesis of disparity analysis either through Foote's pioneering work [Foote Crinoids] or Raup's morphospaces [Raup].
This type of disparity analysis often consists in a more descriptive multidimensionality analysis of the patterns in the diversity of morphological traits and has been successfully used to answer questions such as why are some morphological traits combinations more common than other [Raup] and what are the properties of the resulting morphospace [Gerber].
It is important to highlight that the morphological data used in this context was collected specifically for these analyses and not ``recycled'' from phylogenetic analysis.

\item{Phylogenetic disparity.}
Another disparity approach is to use the same morphological disparity approach as mentioned above but using discrete morphological data commonly used for phylogenetic analysis.
This approach is widely used in palaeobiology to answer a range of macroevolutionary questions [CITEs string].
Although very useful for describing macroevolutionary patterns, this approach can tend to artificially increase disparity between phylogenetically distinct groups due to the nature of so phylogenetic characters use to discriminate groups [CITE Foote].

\item{Functional disparity.}
Functional disparity has since then been assumed to work at a large scale and can be used to study convergence (species with similar trait combination can assume similar functions).
Furthermore there as been work on linking form and function (Diaz paper).
However, the link between trait and function can sometimes be hard to grasp since traits can be linked to multiple functions and multiple functions can be linked to a single trait.

\item{Ecological disparity.}
Linking to functional disparity, people have proposed to use morphological trait disparity as a proxy for ecological niches [CITE].
This seems to work at a macro-evolutionary scale but is still debated at a micro-evolutionary scale [CITE plant trait paper].
For example, the concept of different ecological guilds can easily be distinguished through the fossil record (apex predator have share morphological characteristics through time; e.g. large body mass, pointy teeth, etc.) but can be more difficult at a finer scales and when defining the actual ecological niche (i.e. does a T-Rex occupy the same top predator niche as an Orca?).

\end{enumerate}

This diversity of patterns that can be studied using disparity analyses links to the ambiguous distinction between disparity as a pattern or a process.
In other words, do these disparities represent anything more than the variance of those traits or is it link to specific biological properties of these traits?
For example, does the distribution of traits in apex predators allow them to be distinguished in disparity analysis (i.e. disparity is the pattern) or is there something about being an apex predator that results in this combination of traits (i.e. disparity is the process).
As highlighted above, for the rest of this review, we will essentially use disparity as a term for describing morphological traits patterns but we acknowledge that disparity can sometimes also be the process leading to the observed patterns (e.g. in the case of trait convergence).
Furthermore, as a best practice, we do advice researchers to select the traits that will be fit there question \textit{a priori} and not measure all the traits available.
For example, when studying disparity in locomotion in a group of animals, postcranial characters might more readily answer the question than cranial ones.

\section{Trait disparity}
% IN -> If it's just traits, can all of the traits be used?
In essence, disparity studies are multidimensional analyses of morphological traits.
Such traits can be collected using three main traits: 1) discrete ``cladistic'' characters (e.g. coded as the absence or presence of features [CITE any paleo disp paper]); 2) continuous measurements (e.g. length of features [CITE Anderson fish paper]) or 3) more sophisticated mathematical measurements from in geometric morphometric data (e.g. Pcrocrustes superimpositions of entire features [CITE a geomorph paper], Fourier contour analyses [CITE], etc.).

None of these types of data are fundamentally better and each have a range of materialistic pros and cons (e.g. cost, time, range, etc.), however, some specific questions or methodological pipelines can be more appropriated to some types of data.
For example, when investigating variation within bat wing shapes, both homologous landmarks (e.g. used in geometric morphometric analysis) or continuous measurements of bones may be appropriate to capture patterns of variations.
Conversely, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question from a mechanical perspective.
In the above example, if the question is about functional disparity, say, whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g. wingspan, aspect ratio, etc.).
On the opposite, if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g. digit length, integumentary system, etc.).
It is therefore important to consider disparity patterns as traits disparity patterns approximating biologically complex patterns.

\subsection{Are different traits equivalent?}

From the example above, the question can be asked whether using different methods to approximate the same traits on the same features would lead to similar results.
In other words, does discrete, continuous or geometric morphometric data collected from a same feature capture the same disparity signal?
This question has been assessed several times by comparing the ``disparity signal'' between cladistic and geometric morphometric data in vertebrates leading to the conclusions that data originating from the same source leads to the same results [CITE].
Although the advantages and caveats of each type of traits are well known, there is no clear boundaries to which extent the equivalence between dataset holds.
These equivalence might be more primordial at the trait level (i.e. which traits do we measure) rather than at the measurement method level (i.e. how do we measure them) [Foote Crinoids].

\subsection{Does the part represent the whole?}

In this context, it can be important to understand the relations between the analysis of one specific set of traits and the macroevolutionary questions asked at a organismal level.
For example, when studying mammalian morphological evolution through time based on cladistic characters, do the disparity patterns reflect the changes through time of mammalian morphological niches or do they merely approximate variation in one specific set of traits?
In mammalian discrete datasets, many characters are dental and may not actually represent one organsims' niche (i.e. if the dataset is missing other crucial characters like locomotory ones). %TG: maybe add an example with stuff with the same dentition but living in a different environement, e.g. trees/ground).
However, studies have shown that
-[mammalian teeth are not that bad - Zou and Zhang] 
-[the head can be enough to represent disparity - Melanie's work]

Furthermore, trait data suffers from the same short comings as most data in biology (i.e. allometry, homology, sexual dimorphism, etc...) or to short comings from data in general (missing, non-overlapping, hierarchical - e.g. inapplicable -, ambiguous, polymorphic or correlated, sample size [String of morpho data problems cite]).
Ultimately, data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
% OU -> And how do these traits need to be analysed?

\section{Disparity is multidimensional}

Additionally to pitfalls and caveats stemming from the biological questions and the physical datasets analysed through the lens of morphological disparity, the multidimensional aspect of these analysis can also add a layer of technical complexity.
In fact, disparity analysis are inherently multidimensional and our human brains and the physical word is bound to three spatial dimensions.
This problem is sometime apparent in over-interpretation of disparity analysis results.
For example, in a classic morphological disparity pipelined analysis using geometric morphometric data, the collected data's dimensionality is the number of collected landmarks and their 3 or 2 dimensional set of coordinates.
However, using dimensionality reduction techniques can lead to misinterpretation of the results if not careful.
In the case of ordination, it is not uncommon to then use only the first axis of ordination (e.g. PC1) and draw conclusions based on this axis.
The changes in values along this dimension can display some disparity pattern or gradient at a glance but don't often reflect the actual changes of the set of all the landmarks' dimensions [CITE Vera and Thomas paper].
Therefore, whether for visualising or analysing disparity data, it is crucial to remember it's multidimensionality aspect.

% IN -> Disparity is multidimensional by essence
% Visualisation problem
\subsection{Visualising the multidimensional problem}
\label{visualisation}
Visual interpretations of multidimensional data can easily be misleading.
For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
Conversely, when two groups do not overlap on a two or three dimensional plot, this does not mean they never overlap in the multidimensional space.
Furthermore, multidimensional spaces are not necessary Euclidean.
Such non-Euclidean spaces often have non-intuitive properties, for example straight lines in some dimensions are not straight in the space and, even less intuitively, the distance between points A and B might not be equal to the distance between points B and A [CITE Sylvain].

@@

\subsection{To ordinate or not to ordinate; that is the (multidimensional) question?}
Similarly as for visualisation, reducing the dimensionality of a disparity dataset can be tricky.
Ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g. selecting only the axes that describe up to 95\% of the variation).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions.
This has clear advantages for interpreting the results.
For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).

@@

Because of these points, we strongly recommend that data for multidimensional analyses is not automatically ordinated, and that careful consideration is given to whether the question can be answered without ordination.

%     What do we measure from these multidimensional spaces?
\subsection{Measuring disparity}

A second important point stemming from the multidimensional nature of disparity analyses is considering how to extract some meaningful interpretable summary information: the disparity metric (or index).
As with any summary metric, and because of this reduction in dimensionality, disparity metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
It is often beneficial to use more than one metric to summarise the different aspects of interest of the morphospace (e.g. its width or spread).

The choice of disparity metric is thus essential to the disparity analysis and should always be driven by the question.
When considering only one dimension, disparity metrics can be used to reflect the ``size'' of the distribution (e.g. the range, quantiles or variance), or reflect the most likely value (the central tendency, i.e. the mean, median or mode).
Among these metrics, some will have more preferable properties than others, such as sensitivity to outliers (range, mean or mode are highly sensitive \textit{vs.} quantiles, variance or median are less sensitive) and will thus be more or less appropriate to the different questions and the properties of the distribution.
For example, if the question refers to ``size'' of a group in the morphospace (e.g. ``does group A occupy as much space as group B?''), metrics relating to the spread of the group in the morphospace are the most appropriate (e.g. volume [CITE], variance [CITE], range [CITE], distance from the centroid [CITE]).
Conversely, if the question refers to the ``position'' of a group in the morphospace (e.g. ``does group A occupy the same space as group B?''), metrics relating to the distance between the elements within a group and a fixed point in the morphospace are most appropriate.
Finally, one might also be interested in the density of elements in some regions of the morphospace (e.g. ``is group A more dense than group B?''), metrics relating to the pairwise distances between elements are most appropriate (e.g. nearest neighbour distance, pairwise distances, etc.).

Additionally to the desired properties of what the metrics should capture, it is important to also keep in mind the mathematical properties of the metrics and their associated caveats.
For example, measuring the full sum of the variance of each dimension of the space does not require one to add the co-variance between the axes in a ordinated space using a PCA.
This is not true however in other mathematical spaces or (even in a PCA) when not all dimensions or elements are considered (e.g. when only looking at a group within the morphospace, the sum of variances is not mathematically equal to the sum of the variance of each dimension in the space).
Furthermore, multidimensional space has some counter-intuitive properties that need to be taken into account such as the curse of dimensionality [CITE].
For example, product-based metrics as proxies of volumes (e.g. product of ranges, hypervolume, hypercube, etc.) will tend towards zero fairly quickly for spaces with even a modest number of dimensions [CITING Bellman 1957; also see \url{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners} for an excellent illustration of the problem and Donoho 2000 Aide-Memoire].
Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size, for example range [CITE] or convex hull based metrics [CITE].

% OU -> And how do we test these things?
The technical properties of multidimensional disparity analyses themselves allow for the vast amount of methods and metrics described above to be used in specific context.
However, as mentioned above equally, these tools should always be used in the frame of a specific stated scientific question.
This leads to the importance of clearly defining and rigorously testing on or more hypotheses.

\section{Disparity hypotheses}
% IN -> So how do we go for testing hypotheses in disparity?
%     What are our null expectations?
The multidimensional statistical toolkit has been greatly developed in the last years [CITE Dean Adams, Simone, etc.] but oddly enough not routinely applied to disparity analyses.
Instead, disparity hypotheses testing seemed to be confined to a small set of (well established) methods.
For example, one commonly used test in morphological analysis, is the non parametric permutation analysis of variance [NPANOVA or PERMANOVA; CITE Anderson + examples].
This test is an analysis of variance in the pairwise distances between different groups and is often completely independent of the type of morphospace and the disparity metric used.
Although perfectly valid, this test can sometimes not be directly related to the hypotheses stated.
For example, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the question.

@@

\subsection{Phylogeny}
More specifically within the realm of testing hypothesis, it is important to remember that when analysing biological data, the data is not independent [CITE Felsenstein or so].
For many disparity analyses of macroevolutionary questions, phylogeny must thus be taken into account, whether for including ancestral state estimates or to incorporate phylogenetic structure in the data.
For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
and therefore will not covered further in this review.

\subsubsection{Ancestral state estimation}
When looking at disparity in the past (e.g. disparity-through-time) ancestral state estimations are a common approach.
Ancestral state estimation can be performed in two ways:
\begin{enumerate}
\item pre-ordination: the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original dataset[CITE].
\item post-ordination: the estimation is done after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
\end{enumerate}
Pre-ordination ancestral state estimation can change the ordinated space's geometry (i.e. the relationship between the points not estimated) and implies longer computational times but, once the morphospace is defined, it will remain constant (i.e. its properties will not change).
Post ordination ancestral state estimation will not change the ordinated space's geometry and will be faster to compute, however, it will change the morphospace's properties such as the number of parameters (i.e. elements in the space - this can be problematic for statistical tests down the line) or the variance on each axis [CITE Graeme's paper].

@@

Alternatively, ancestral state estimation can be used to provide realistic null expectations from the data.
In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes [CITE time slice].
Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution [CITE STD paper?].
%     How can we analyse it through time?

@@

\subsection{Conclusion}
However, as with any great power, it comes with great responsibility. 
With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.
